---
title: "Assignment 04"
author: "Sahana Srinivasan and Maria Antonella Pareira" 
output: html_document
urlcolor: blue
---

## Visualizing Changes in Labor Market Outcomes for Women in the United States
This document identifies trends in labor market outcomes for women in the United States using data from the Current Population Survey (CPS) from 1980 to 2020. The purpose of these visualizations is to identify potential changes in employment and labor market participation that could be explained by the onset of the COVID-19 pandemic with further research. 

Data files and the data dictionary for the CPS can be found [at this link](https://cps.ipums.org/cps/about.shtml). 

### Data Cleaning and Setting Up
```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(srvyr)
tinytex::install_tinytex()
```

```{r, echo = FALSE}
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

```{r, echo = FALSE}
#Data Cleaning and Processing
library("ipumsr")
library("tidyverse")
library("dplyr")
ddi <- read_ipums_ddi("cps_00002.xml")
data <- read_ipums_micro(ddi)
data1 <- data %>% 
  drop_na(WTFINL)
library(srvyr)
cps_svy <- as_survey_design(data1, weights = WTFINL)
class(cps_svy)
class(data)
```

## **Figure 1: Female Employment Rate (1980 - 2020)**

Figure 1 shows the change in female employment rates in the United States from 1980 to 2020 at the start of each decade. From 1980 to 2000, female employment rate increased from X to X. The employment rate fell sharply by 2010 to X and continued to decline to X by 2020. Further analysis could explore whether these drops in employment among women are related to specific shocks, such as the 2008 real estate crisis or the start of the COVID-19 pandemic in late 2019.

```{r}
#Figure 1
library("ggplot2")
library("haven")
cps_svy %>% 
  mutate(female = if_else(SEX == 2, 1, 0)) %>% 
  mutate(employed = if_else(EMPSTAT == 10 | EMPSTAT == 12, 1, 0)) %>%
  mutate(unemployed = if_else(EMPSTAT == 10 | EMPSTAT == 12, 0, 1)) %>%
  filter (female == 1,  AGE >= 16) %>% 
  group_by(YEAR) %>% 
  summarize(
    employed = survey_mean(employed == 1)
  ) %>% 
  ggplot(mapping = aes(x = YEAR, y = employed)) + 
  geom_line() + 
  geom_point() + 
  theme_minimal() + 
  labs(
    title = "Figure 1. Female Employment Rate 1980-2020, Upon Start of Each Decade",
    subtitle = "Female employment rate peaked in 2000 and has decreased since", 
    x = "Year", 
    y = "Female Employment Rate", 
    caption = "Source = Sarah Flood, Miriam
    King, Renae Rodgers, Steven Ruggles, J. Robert
    Warren and Michael Westberry. Integrated Public
    Use Microdata Series, Current Population Survey:
    Version 9.0 [dataset]. Minneapolis, MN: IPUMS,
    2021. https://doi.org/10.18128/D030.V9.0*"
    )
```

## **Figure 2: Female Employment by Generation**

Figure 2 shows employment rates by generation for women in the United States in 2020. Unsurprisingly, employment is lowest among women who belong to the “Post War” generation (ages 75 to 92) and the “Boomers 1” generation (ages 74 to 66). Employment is highest among Millennial (ages 24 to 31) and Gen X (40 to 55). Employment is much lower among Gen Z women, that is, women aged 16 to 23; future analysis could evaluate whether lower employment among this cohort is related to the impacts of the COVID-19 pandemic on recent high school and college graduates’ ability to enter the labor market. 

```{r}
#Figure 2 
cps_svy %>% 
  mutate(female = if_else(SEX == 2, 1, 0)) %>% 
  mutate(age_category = case_when( #create variable for generations 
    AGE <= 9 ~ "Youth",
    AGE <= 25 ~ "Gen Z" , 
    AGE <= 41 ~ "Millenials", 
    AGE <= 57 ~ "Gen X", 
    AGE <= 67 ~ "Boomers 2", 
    AGE <= 76 ~ "Boomers 1",
    AGE <= 94 ~ "Post War",
    TRUE ~ "WWII"
        )
    ) %>% 
  mutate(
    employed = if_else(EMPSTAT == 10 | EMPSTAT == 12, 1, 0)) %>% 
  filter(YEAR == 2020) %>% 
  filter(female == 1, AGE >= 16) %>%
  group_by(age_category) %>%
  summarize( 
    employed = mean(employed == 1), 
  ) %>% 
ggplot(mapping = aes(x = age_category, y = employed)) + 
geom_col(color = "blue", fill = "red") + 
  scale_x_discrete(limits = c("Post War", "Boomers 1", "Boomers 2", "Gen X", "Millenials", "Gen Z")) +
  theme_minimal() + 
   labs(
    title = "Figure 2. Proportion of Females Employed By Generation in 2020", 
    subtitle = "Female employment is highest among millenials and lowest among post-war", 
    x = "Generations", 
    y = "Proportion of Females Employed", 
    caption = "Source = Sarah Flood, Miriam
    King, Renae Rodgers, Steven Ruggles, J. Robert
    Warren and Michael Westberry. Integrated Public
    Use Microdata Series, Current Population Survey:
    Version 9.0 [dataset]. Minneapolis, MN: IPUMS,
    2021. https://doi.org/10.18128/D030.V9.0*"
    )
```

## **Figure 3: Average Age of Women not in the Labor Force by Race in 2010 and 2020**

Figure 3 shows the distribution of ages for women not in the labor force disaggregated by race in 2010 and 2020. The average age of non-employed women was higher in 2020 than in 2010 for women of all races. White and Black non-employed women have higher average ages than non-employed women from all other races in both years. Further research should evaluate whether the COVID-19 pandemic exacerbated economic conditions enough that women were largely incentivized to delay the onset of their retirement, thereby increasing the average age of women not in the labor force. Also, further analysis should take into account household income and asset levels for women not in the labor force across all races.  

```{r}
#Figure 3: 
ggplot1<- data1%>%  
mutate(race = case_when(
    RACE == 100 ~ "White",
    RACE == 200 ~ "Black", 
    RACE == 300 ~ "American Indian/Aleut/Eskimo", 
    RACE == 650 | RACE == 651 ~ "Asian", 
    RACE <= 830 | RACE >= 700 ~ "Other")) %>%
  mutate(
    labor_force = if_else(LABFORCE == 2, 1, 0)) %>% 
filter (YEAR == 2010) %>%
filter (SEX == 2, AGE >= 16, labor_force == 0) %>%
ggplot(mapping = aes( x = race , y = AGE)) + 
geom_boxplot() + 
  theme_minimal() + 
  theme(text=element_text(family="ArialMT", size=7)) +
   labs(
    title = "Figure 3. Average Age of Women not in the Labor Force by Race in 2010 and 2020", 
    subtitle = "2010",
    x = "Race", 
    y = "Age")


library("patchwork")
ggplot2 <- data1%>%  
mutate(race = case_when(
    RACE == 100 ~ "White",
    RACE == 200 ~ "Black", 
    RACE == 300 ~ "American Indian/Aleut/Eskimo", 
    RACE == 650 | RACE == 651 ~ "Asian", 
    RACE <= 830 | RACE >= 700 ~ "Other")) %>%
 mutate(
    labor_force = if_else(LABFORCE == 2, 1, 0)) %>% 
filter (YEAR == 2020) %>%
filter (SEX == 2, AGE >= 16, labor_force == 0) %>%
ggplot(mapping = aes( x = race , y = AGE)) + 
geom_boxplot() + 
  theme_minimal() + 
  theme(text=element_text(family="ArialMT", size=7)) +
   labs(
    subtitle = "2020",
    x = "Race", 
    y = "Age", 
    caption = "Source = Sarah Flood, Miriam
    King, Renae Rodgers, Steven Ruggles, J. Robert
    Warren and Michael Westberry. Integrated Public
    Use Microdata Series, Current Population Survey:
    Version 9.0 [dataset]. Minneapolis, MN: IPUMS,
    2021. https://doi.org/10.18128/D030.V9.0*"
    )

ggplot1 / ggplot2 
  


```

## **Figure 4: Employment and Non-employment Numbers by Highest Level of Education Completed (2020)**

This graph depicts numbers of employment and non-employment disaggregated by highest level of education attained by 2020. Though a noticeably greater share of those with an elementary or middle school education are non-employed than are employed, the opposite is true among those whose highest level of education is Bachelor’s or Graduate; rates are comparable for those with a high school education. Further analysis should explore the extent to which the COVID-19 pandemic reduced employment opportunities for those with an elementary or middle school education and, conversely, allowed for the retainment of employment opportunities for individuals with a post-secondary degree. Studying the type of employment opportunities lost or gained in 2020 can also elucidate these findings.


```{r}
#Visualization 4 - 
data1 %>% 
  mutate(employed = if_else(EMPSTAT == 10 | EMPSTAT == 12, "Employed", "Not Employed")) %>% 
  mutate(ed_level = case_when( 
    EDUC < 032 & EDUC >= 021 ~ "Elementary", 
    EDUC < 070 ~ "Middle", 
    EDUC < 100 ~ "High", 
    EDUC < 122 ~ "Bachelors", 
    EDUC <= 125 ~ "Graduate" 
    )) %>% 
  filter(SEX == 2, AGE >= 16, YEAR == 2020) %>% 
  ggplot(mapping = aes(x = ed_level, y = employed), position = "dodge") + 
  geom_count() + 
  scale_size_area(max_size = 30) + 
  scale_x_discrete(limits = c("Elementary", "Middle", "High", "Bachelors", "Graduate")) +
  ggtitle("Figure 4. Employment and Non-employment Numbers by Highest Level of Education Completed (2020)") + 
  theme_minimal() +
  theme(text=element_text(family="ArialMT", size=7)) + 
   labs(
    subtitle = "Employment is larger than non-employment for women with bachelors and graduate degrees but not for lower levels of education",
    x = "Highest Level of Education Completed by 2020", 
    y = "Employment Status", 
    caption = "Source = Sarah Flood, Miriam
    King, Renae Rodgers, Steven Ruggles, J. Robert
    Warren and Michael Westberry. Integrated Public
    Use Microdata Series, Current Population Survey:
    Version 9.0 [dataset]. Minneapolis, MN: IPUMS,
    2021. https://doi.org/10.18128/D030.V9.0*"
    ) 
```